Simulating High-Dimensional Multivariate Data

### 2020-10-29 030-bigsimr-package.Rmd
bigsimr R package rev2

The `bigsimr` package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
When designing `bigsimr`, we aimed to conveniently provide parallelized computation, with multicore and GPU acceleration.

This section describes basic of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
This workflow proceeds from data, to estimation, simulation configuration, and random vector generation.
For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outways the computation gains. 
Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
The section concludes with a description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package.

@@

## Basic use workflow

We're going to show the basic use and syntax of `bigsimr` by using the New York air quality data set (`airquality`) included in the R `datasets` package.
First we load the `bigsimr` library and a few other conveninent data science packages, including the syntaxically-pleasing `tidyverse` suite of R packages.

@@

For simplicity and to provide a minimal worknig example, we'll consider bivariate simulation of temperature, in degrees Fahrenheit, and ozone level, in parts per billion.

@@

Next we must specify the marginal distributions and correlation coefficient (type and magnitude).
Here the analyst is free to be creative and we will not take up goodness-of-fit considerations.
But it seems sensible without domain knowledge to estimate these quantative from the data and we have a suite of functions designed for this task.

*Specifying marginal distributions*

Though we don't know the true distribution of ozone levels or air temperatures, we'll assume that both are log-normally distributed since the values are positive and skewed.
We'll use the well-known maximum likelihood estimators for the lognormal parameters:

@@

Since our estimated correlation is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
By comparison, the Pearson correlation is restricted for these margins.

@@

### Simulating Multivariate Distributions

Let's now simulate 10,000 observations from the joint distribution using `bigsimr::rvec`

```{r ch030-sim-margins, echo=TRUE}
x <- rvec(10000, rho, margins, "pearson")

df_sim <- as.data.frame(x)
colnames(df_sim) <- colnames(df)
```

and visualize the joint distribution

```{r ch030-plot-sim}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Temperature", y = "Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

@@

### Comparison to Uncorrelated Samples

We can compare the bivariate distribution above to one where no correlation is taken into account.

```{r ch030-compare-uncorrelated}
df_sim2 <- data.frame(
  Temp = rnorm(10000, mean(df$Temp), sd(df$Temp)),
  Ozone = rlnorm(10000, mle_mean(df$Ozone), mle_sd(df$Ozone))
)

@@

@@

### Creating a custom marginal distribution {#custom-margins}

Because `bigsimr` uses an `alist` to store the margins, any kind of probability distribution can be used including custom marginal distributions not provided in base R. This accommodation is possible because the main function `rvec` uses the inverse CDF of a margin to transform from correlated uniform margins to the target distribution.

It is highly recommended that users follow R's naming convention for probability distributions - I.e. prefixing distributions with `r` and `q` for *random* and *quantile* respectively. Only the quantile function is necessary to generate multivariate data, but if one wishes to compute the theoretical correlation bounds, then one should also specify the random number generator (RNG) function for their custom distribution.

For an example, R does not have any functions for the *Pareto* distribution. We can write a quantile function and RNG function and use it in `rvec`.

The Pareto CDF is defined as

@@

Next we define the Pareto quantile function in R. Note that extra checks should be taken to ensure that the arguments are valid. We omit such checks for clarity.

@@

Writing a RNG function for our new distribution can be accomplished by calling the quantile function on the output of a random unit uniform variable.

@@

### Calculating the nearest positive (semi)definite correlation matrix

### Using multicore

### Utilizing the GPU

The `bigsimr` packages uses Google's `jax` python library to interface with an Nvidia CUDA enabled GPU.

### Simulation-based computation of correlation bounds

### Using `bigsimr` on a computing cluster via `rslurm`

Though `bigsimr` runs quickly, at large $d$ users may want to run jobs on a shared computing server. The R package `rslurm` makes it easy to run embarrassingly large parallel `rvec` calls. This example assumes that `bigsimr` is installed on a system with a slurm scheduler installed.

```{r ch030-rvecRslurm, echo=FALSE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_single_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
sjob <- slurm_call(rvec, jobname = 'rvec',
                   list(n=1e6,
                        rho = rho,
                        margins = margins,
                        type = "spearman"),
                   submit = TRUE)
```

```{r ch030-rvecRslurmRes, echo=FALSE, eval=FALSE}
res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
head(res, 3)
## cleanup_files(sjob) ## careful this deletes output also
```

Now, let's show off the real power of combining `bigsimr` and `rslurm` by simulating many correlation structures for these three marginals. The `rslurm::slumr_map` syntax mirrors the familiar `base::lapply` and `purrr::map` functions.

```{r ch030-rvecRslurmMap, echo=FALSE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
set.seed(06202020)
simReps <- 100
rhoList <- replicate( n = simReps, bigsimr::rcor(d = length(margins), constant_rho = TRUE), simplify=FALSE )
sjob <- slurm_map(x = rhoList,
                  f = rvec,
                  jobname = 'rvecMap',
                  n=1e6,
                  margins = margins,
                  type = "spearman",
                  nodes = 4,
                  cpus_per_node = 4,
                  submit = TRUE)

@@

On a cluster carrying 24 nodes with 48 threads, these 100 jobs completed in about a minute. Let's evaluate the quality of simulation performance.

```{r ch030-rvecRslurmMapRes, echo=FALSE, eval=FALSE}
list.files('_rslurm_rvecMap', 'results')
res <- rslurm::get_slurm_out(sjob, outtype = 'raw')
rhoHats <- lapply( res, bigsimr::fastCor, method = 'spearman' )

resData <- NULL
simreps <- length(rhoHats)
for (i in 1:simreps){
    tmpRhoHat <- rhoHats[[i]]
    ## average together
    margins
    avgRhoHat <- tmpRhoHat[lower.tri(tmpRhoHat)]
    tmpPair <- paste0( "pair", apply( combn(x = length(margins), 2), 2, paste, collapse = "_" ) )
    tmpResData <- data.frame( simNum=i, rho=rhoList[[i]][1,2], rhoHat=tmpRhoHat[lower.tri(tmpRhoHat)], pair=tmpPair )
    resData <- rbind(resData, tmpResData)
}

## "wrangle" into a tibble
library(tidyverse)
res <- as_tibble(resData)
## save for viz
saveRDS( res, file = 'data/rvecRslurmMapRes.rds' )

## cleanup careful this deletes output also
## cleanup_files(sjob) 
```

```{r ch030-rvecRslurmMapFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias. Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}
library(tidyverse)

res <- readRDS( 'data/rvecRslurmMapRes.rds' )
res

res %>%
    ggplot( aes( x = rho, y = rhoHat, color = pair ) )  +
    geom_point() +
    geom_abline( slope = 1, intercept = 0 )
```

>>>>

The `bigsimr` package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms) for details).
When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration.

This section describes basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result vizualization.
For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outways the computational gains. 
Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package.

@@

## Basic use illustrated through a minimal example

We'll demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package.
First, we load the `bigsimr` library and a few other convenient data science packages, including the syntacically-pleasing `tidyverse` suite of `R` packages.

@@

For simplicity and to provide a minimal working example, we'll consider bivariate simulation of temperature, in degrees Fahrenheit, and ozone level, in parts per billion.

@@

Next we must specify the marginal distributions and correlation coefficient (type and magnitude).
Here the analyst is free to be creative.
For this example, we will not take up goodness-of-fit considerations.
But it seems sensible without domain knowledge to estimate these quantities from the data and we have a suite of functions designed for this task.

*Specifying marginal distributions*.
Though we don't know the true distribution of ozone levels or air temperatures, we'll assume that both are log-normally distributed since the values are positive and skewed.
We'll use the well-known maximum likelihood estimators for the lognormal parameters:

@@

Since our estimated Spearm correlation $\hat{ \rho_S }$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
By comparison, the Pearson correlation $\rho_P$ is restricted to [-0.8648, 0.8727] for these margins (see `bigsimr::cor_bounds`output below).

@@

*Simulating random vectors*.
Finally, we arrive at the main function of `bigsimr`, `rvec`.
Let's now simulate $B=10,000$ observations from the assumed joint distribution of Ozone levels and Temp.

```{r ch030-sim-margins, echo=TRUE}
x <- rvec(10000, rho, margins, type)
df_sim <- as.data.frame(x)
colnames(df_sim) <- colnames(df)
```

Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.

```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimick the observed data with respect to both the marginal characteristics and bivariate association."}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Temperature", y = "Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

@@

<!-- 
*Comparison to Uncorrelated Samples*.
We can compare the bivariate distribution above to one where no correlation is taken into account.

```{r ch030-compare-uncorrelated, eval=FALSE}
df_sim2 <- data.frame(
  Temp = rnorm(10000, mean(df$Temp), sd(df$Temp)),
  Ozone = rlnorm(10000, mle_mean(df$Ozone), mle_sd(df$Ozone))
)

@@

your comment -->

@@

*Creating a custom marginal distribution*.
Because `bigsimr` uses an `alist` to store the margins, any kind of probability distribution can be used including custom marginal distributions not provided in base `R`. 
This accommodation is possible because the main function `rvec` uses the inverse CDF of a margin to transform from correlated uniform margins to the target distribution.
Therefore, to specify a marginal distribution absent from an existing packge, the user simply needs to provide a closed form expression from the corresponding quantile function.

It is important to follow `R`'s naming convention for probability distributions.
Users should prefix distributions with `r` and `q` for *random* and *quantile* respectively.
Only the quantile function is necessary to simulate random vectors, but to compute the theoretical correlation bounds, it is required to supply univariate random number generator as well.

For an example, let's provide a custom *Pareto* distribution for use with `bigsimr`.
The Pareto CDF is

@@

Next we define the Pareto quantile function in R.

@@

For example, writing random number generating function for our target marginal can be accomplished by calling the quantile function on a uniformly distributed random variable (the inverse transform method @Rizzo2007).

@@

*Using `bigsimr` on a computing cluster via `rslurm`*.
Though `bigsimr` runs quickly, at large $d$ users may want to run jobs on a shared computing server.
The R package `rslurm` makes it easy to run embarrassingly large parallel `rvec` calls.
This example assumes that `bigsimr` is installed on a system with a slurm scheduler installed.

```{r ch030-rvecRslurm-with-comments, echo=FALSE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_single_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
sjob <- slurm_call(rvec, jobname = 'rvec',
                   list(n=1e6,
                        rho = rho,
                        margins = margins,
                        type = "spearman"),
                   submit = TRUE)
```

```{r ch030-rvecRslurm, echo=TRUE, eval=FALSE}
library(rslurm)
sjob <- slurm_call(rvec, jobname = 'rvec',
                   alist(n=1e6,
                         rho = rho,
                        margins = margins,
                        type = "spearman"),
                   submit = TRUE)
```

<!-- 
Once your jobs complete, `rslurm` conveniently stores simulation results in a standardized format.
Retriving the data is accomplished by
-->

```{r ch030-rvecRslurmRes, echo=FALSE, eval=FALSE}
res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
head(res, 3)
```

Now, let's show off the real power of combining `bigsimr` and `rslurm` by simulating many correlation structures.
The `rslurm::slumr_map` syntax mirrors the familiar `base::lapply` and `purrr::map` functions.

```{r ch030-rvecRslurmMap-comments, echo=FALSE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
set.seed(06202020)
simReps <- 100
rhoList <- replicate( n = simReps, bigsimr::cor_randPSD(d = length(margins)),
                     simplify=FALSE )
sjob <- slurm_map(x = rhoList,
                  f = rvec,
                  jobname = 'rvecMap',
                  n=1e6,
                  margins = margins,
                  type = "spearman",
                  nodes = 4,
                  cpus_per_node = 4,
                  submit = TRUE)

@@

```{r ch030-rvecRslurmMap, echo=TRUE, eval=FALSE}
library(rslurm)
simReps <- 100
rhoList <- replicate( n = simReps, bigsimr::cor_randPSD(d = length(margins)), simplify=FALSE )
sjob <- slurm_map(x = rhoList,
                  f = rvec,
                  jobname = 'rvecMap',
                  n=1e6,
                  margins = margins,
                  type = "spearman",
                  nodes = 4,
                  cpus_per_node = 4,
                  cores = 4,
                  submit = TRUE)
```

On a cluster carrying 24 nodes with 48 threads, these 100 jobs completed in about a minute.

<!-- 
Let's evaluate the quality of simulation performance.

your comment -->

```{r ch030-rvecRslurmMapRes, echo=FALSE, eval=FALSE}
list.files('_rslurm_rvecMap', 'results')
res <- rslurm::get_slurm_out(sjob, outtype = 'raw')
rhoHats <- lapply( res, bigsimr::fastCor, method = 'spearman' )
resData <- NULL
simreps <- length(rhoHats)
for (i in 1:simreps){
    tmpRhoHat <- rhoHats[[i]]
    ## average together
    margins
    avgRhoHat <- tmpRhoHat[lower.tri(tmpRhoHat)]
    tmpPair <- paste0( "pair", apply( combn(x = length(margins), 2), 2, paste, collapse = "_" ) )
    tmpResData <- data.frame( simNum=i, rho=rhoList[[i]][1,2], rhoHat=tmpRhoHat[lower.tri(tmpRhoHat)], pair=tmpPair )
    resData <- rbind(resData, tmpResData)
}
## "wrangle" into a tibble
library(tidyverse)
res <- as_tibble(resData)
## save for viz
saveRDS( res, file = 'data/rvecRslurmMapRes.rds' )
## cleanup careful this deletes output also
## cleanup_files(sjob) 
```

```{r ch030-rvecRslurmMapFig, echo = FALSE, eval=FALSE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias. Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}
res <- readRDS( 'data/rvecRslurmMapRes.rds' )
res
res %>%
    ggplot( aes( x = rho, y = rhoHat, color = pair ) )  +
    geom_point() +
    geom_abline( slope = 1, intercept = 0 )
```

### 2020-10-29 060-conclusion-discussion.Rmd
migrating ESCO 2020 talk into sections, revising everything
Revising background
bigsimr R package rev1
bigsimr R package rev2

# Conclusion/Discussion

@@

@@

Mention `rslurm`

>>>>

# Conclusion and discussion {#discussion}

@@

-  Simulating correlated discrete and count data takes more care [@Xia17], including slightly redefining the inverse. 
- a user can always supply their own input Pearson after using the existing matching schemes [@XZ19].

@@

- mention that `bigsimr`

Nvidia-based(copyright symbol) graphical-processing-unit (GPU), OpenCL future direction.

### 2020-10-29 070-supplementary-materials.Rmd
Revising background
bigsimr R package rev2

# Supplementary Materials

We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github.

@@

The authors gratefully acknowledge the helpful discussions with Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.

@@

The authors report no conflict of interest. DO WE?

@@

FUNDING FROM EVERYONE HERE. CP3 grant.

# Nomenclature/Notation

>>>>

# Supplementary Materials {#misc}

We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github, https://schisslergroup.github.io/bigsimr/.

@@

The authors gratefully acknowledge the helpful discussions with University of Arizona's Professor Walter W. Piegorsch and Professor Edward J. Bedrick during this project's conception.
We also gratefully acknowledge Heather Knudson's graphic design for the `bigsimr` R Package.
The results published here are in whole or part based upon data generated by the TCGA Research Network: https://www.cancer.gov/tcga.

@@

The authors report no conflict of interest.

@@

Research reported in this publication was supported by MW-CTR-IN of the National Institutes of Health under award number NIH 1U54GM104944.

<!--

Nomenclature/Notation
